\emph{icepack}: a new glacier flow modeling package in Python, version 1.0

### 2021-04-28 reviews/response-3.tex

\begin{document}

\title{Response to review \#3}
\author{}
\date{}

\maketitle

Thank you for this thorough review!
Original comments are in blue, our responses are in black, and hyperlinks in magenta.

@@

We do not claim to have reduced the need for expertise in glaciology itself, i.e. understanding mass and energy balance, etc.
We also consider having some understanding of the different boundary conditions that might apply to the system to be part of this knowledge.
The text has been expanded to describe these boundary conditions in more detail (see comments that follow).
Granted, (1) there is more demonstration work that we need to do in order to cover use cases like modeling a small mountain glacier and (2) there are technical hurdles we still need to overcome to do a better job handling the margins.
Very likely your recent work on variational inequalities will factor into improvement on the second front.

We have altered the text to state more clearly that our goal is to help users who are not experts in scientific computing.
We added the following to the introduction: ``We have focused efforts thus far on process studies of individual glaciers or drainage basins (use cases 2 and 3 of the list above).
Development of icepack is ongoing and we will broaden our efforts to encompass more use cases in future.''
We also added the following sentence to the conclusion: ``Exactly how to apply principles from HCI to maximize usability is, nonetheless, not an exact science.
Our implementation may have failed to meet this goal and changes in future versions will be guided by what users find most difficult.''

@@

The technique might be applicable in other areas but we haven't had the chance to try it yet.
No change to the text.

\textcolor{blue}{517--522:  Rich problem-aware preconditioners and solvers.  On board with Firedrake.}

\textcolor{blue}{523, A:  This honest, clear sentence is not at all expected by the reader who remembers lines 15-19 in the Introduction.  Please let the reader know earlier.}

We added the following to the introduction: ``We have focused efforts thus far on process studies of individual glaciers or drainage basins (use cases 2 and 3 of the list above).
Development of icepack is ongoing and we will broaden our efforts to encompass more use cases in future.''

@@

\textcolor{blue}{533--586:  The MISMIP+ experiment is part of the 2008-onward tradition of running ice sheet models in rectangular boxes (MISMIP, EISMINT-HOM, ...).  This is fine in a field where one also has some laboratory fluid models, i.e. actual stuff, which fit in boxes.  But this tradition has a distressing impact on new model development, which is to modularize around the choices one makes between intercomparison-specified boxes and their associated boundary conditions.  (Yes, that is what Icepack looks like right now.)}

We included this assuming that it's a necessity to repeat at least some of the standard benchmarks from the literature in order to publish a new model.
No change to the text.

### 2021-06-11 icepack.tex
Final changes
Final technical corrections

Nearly all glaciologists, from graduate students to senior researchers, need to use numerical models at some point in their career.
Several software packages for glacier flow modeling already exist and are effective in the hands of experts.
We highlight four main uses of glacier models in the literature:
\begin{enumerate}
    \item projecting future glacier extent and estimating the sea-level rise contribution from glacier dynamics \citep{joughin2014marine},
    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood \citep{nick2010physically, werder2013modeling},
    \item estimating unobservable quantities, such as bed friction or rheology, from observational data \citep{vieli2007causes, shapero2016basal}, and
    \item reconstructing what glaciers of the near- or distant-past may have looked like \citep{huybrechts2002sea}.
\end{enumerate}
To accomplish these tasks, modeling tools are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
Many glaciologists receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
In this paper, we introduce a new Python software package for glacier flow modeling called \emph{icepack}.
Our goal is to make a tool that will be both accessible to novices and more productive for experts.

@@

The evolution equations for ice thickness and velocity are necessary for any simulation, both other fields with their own dynamics can be a part of the problem as well.
For example, the rheology and friction are the main coefficients in the diagnostic equation.
These coefficients are functions of other fields that have their own evolution equations.
The rheology is a function of temperature, englacial water content, and damage from crevassing.
Likewise, the friction coefficient can be described in terms of the subglacial water pressure and a roughness factor for the underlying bedrock \citep{cuffey2010physics}.
The diagnostic and prognostic equations can then be supplemented with evolution equations for these fields, for example the heat equation for temperature, or a hydrology model for subglacial water pressure.

@@

The mass transport equation for ice thickness is a free boundary problem, where the free boundary is the contour between ice-covered and ice-free regions \citep{schoof2013ice}.
A naively-implemented prognostic solver could erroneously compute negative thickness values in subsequent timesteps when there is ablation in ice-free regions.
A common and ad-hoc approach to work around this issue is to truncate the thickness at zero at every timestep.
The principled approach is to instead treat the free boundary problem directly as a variational inequality \citep{jouvet2012steady}.
Icepack currently lacks a principled scheme for tracking this free boundary and we instead rely on truncation.
Treating this problem as a variational inequality in icepack will be the subject of future development.
PETSc includes scalable solvers for variational inequalities \citep{bueler2020petsc} that are also available through Firedrake.

>>>>

Nearly all glaciologists, from graduate students to senior researchers, need to use numerical models at some point in their career.
Several software packages for glacier flow modeling already exist and are effective in the hands of experts.
We highlight four main uses of glacier models in the literature:
\begin{enumerate}
    \item projecting future glacier extent and estimating the sea-level rise contribution from glacier dynamics \citep{joughin2014marine},
    \item exploring aspects of glacier physics, such as hydrology and calving, that are not completely understood \citep{nick2010physically, werder2013modeling},
    \item estimating unobservable quantities, such as bed friction or rheology, from observational data \citep{vieli2007causes, shapero2016basal}, and
    \item reconstructing what glaciers of the near- or distant-past may have looked like \citep{huybrechts2002sea}.
\end{enumerate}
To accomplish these tasks, modeling tools are usually written in compiled programming languages such as C, C++, and Fortran for reasons of computational efficiency.
Many glaciologists receive little or no formal programming training, much less in these languages, and are instead self-taught in either Python or MATLAB.
The ubiquity of C, C++, and Fortran in scientific computing can create a barrier to entry for glaciologists who are not experts in high-performance computing.
In this paper, we introduce a new Python software package for glacier flow modeling called \emph{icepack}.
Our goal is to make a tool that will be both accessible to novices and more productive for experts.
We have focused efforts thus far on process studies of individual glaciers or drainage basins (use cases 2 and 3 of the list above).
Development of icepack is ongoing and we will broaden our efforts to encompass more use cases in future.

@@

The evolution equations for ice thickness and velocity are necessary for any simulation, but other fields with their own dynamics can be a part of the problem as well.
For example, the rheology and friction are the main coefficients in the diagnostic equation.
These coefficients are functions of other fields that have their own evolution equations.
The rheology is a function of temperature, englacial water content, and damage from crevassing.
Likewise, the friction coefficient can be described in terms of the subglacial water pressure and a roughness factor for the underlying bedrock \citep{cuffey2010physics}.
The diagnostic and prognostic equations can then be supplemented with evolution equations for these fields, for example the heat equation for temperature, or a hydrology model for subglacial water pressure.

@@

The mass transport equation for ice thickness is a free boundary problem, where the free boundary is the contour between ice-covered and ice-free regions \citep{schoof2013ice}.
Where there is ablation in ice-free regions, a correctly-implemented prognostic solver will compute a negative thickness, which is unphysical and requires correction.
The prognostic solver in icepack truncates the thickness to zero, which approximates the free boundary, where there is ablation in ice-free regions.
An implicit approach would instead treat the free boundary problem directly as a variational inequality \citep{jouvet2012steady}.
Icepack currently lacks such a scheme and this will be the subject of future development.
PETSc includes scalable solvers for variational inequalities \citep{bueler2020petsc} that are also available through Firedrake.
